Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-07-24 paper.tex
Add collapse profile test text

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter
representative of the fluid in the Massive Collapsed Star test
(Sec.~\ref{ssec:test_collapse}) at the radius 50~km.
For comparision with that test we use the LS180 equation of state
\cite{latt1991-nuc_eos}
\footnote{available at \lstinline{stellarcollapse.org}}
in which the equilibrium $\nu_e$ neutrino chemical potential is
$\eta_{\nu_e}=-0.1555$
(with $\eta_{\bar{\nu}_e}=-\eta_{\nu_e}$ and $\eta_{\nu_x}=0$).

@@

Expanding the exponential in Eqn.~\ref{eqn:fae_homogeneous_general},
we see the distribution function takes the limiting values:
\begin{equation}
  f_{\rm AE}(\cos A) = f^{\rm eq}
  \begin{cases}
    \chi s(\cos A) & s \ll \chi^{-1}\,\textrm{optically thin}\\
    1                  & s \gg \chi^{-1}\,\textrm{optically thick}.
  \end{cases}
\end{equation}
This agrees with Fig.~\ref{fig:f_absorption_sphere}, with $\nu_x$
in the optically thin limit and $\nu_e$ in the optically thick limit.

@@

\subsection{Post-Bounce Collapse Profile:
  Testing Scattering}
\label{ssec:test_collapse}
We use a single time snapshot of a supernova model in its accretion phase,
100ms after core bounce.
[Standard radiation transport test described in~\cite{ocon2015-gr1d_with_nu},
and in \cite[App.~E.6]{fouc2015-m1_nsbh},
and in~\cite{abdi2012-monte_carlo}.]
\todo{clarify inelastic scattering and velocity terms}
Plot in Fig.~\ref{fig:avg_eps_collapse}.

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-collapse_profile_avg_eps_vs_cosA}
  % Data:   $DIR/
  % Script: $DIR/plot-compare_spectral_vs_gray-angular_distribution_avg_eps-production.gnu
  % w/  DIR_MC     =$DIR/ResultsSherwood-03/Spectra
  %     DIR_SPEC_SC=$DIR/ResultsBrett-A-03/Test-scat-01
  %     DIR_GRAY_SC=$DIR/ResultsBrett-B-03/Test-scat-01
  %     DIR_GRAY_AB=$DIR/ResultsBrett-B-03/Test-noscat-01
  % and DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20161007-ray_tracing_tests/CollapseProfile/
  \caption{Average energies $\langle\varepsilon\rangle(\cos A)=J(\cos A)/G(\cos A)$
    in the collapse profile test described in Sec.~\ref{ssec:test_collapse}.
    The observer is at 500~km, and the shock at 150~km,
    so that the half-opening angle of the shock is $\cos A=0.954$.
    The four methods depicted are 1) a fiducial Monte Carlo calculation,
    ray tracing using the 2) spectral and 3) gray methods to estimate background
    fields for scattering, and 4) ray tracing neglecting scattering.
  }
  \label{fig:avg_eps_collapse}
\end{figure}

@@

\begin{table}%[h]
  \caption{
    Comparison of total luminosities and average energies between
    ray tracing and the M1 transport simulation.
    The ray tracing totals were computed from sums over
    observers placed in the $y$-$z$ plane along an arc at coordinate distance
    $r=250\,{\rm km}$ from the center of the star, at 30 positions
    distributed evenly in $\cos\theta$ over the colatitude range
    $\theta\in[0,90^{\circ}]$.
    M1 values are taken from \cite[Figs. 7, 9, 10]{fouc2016-m1_evolve_n}.
  }
  \label{tab:nsns_rt_vs_m1}
  \begin{tabularx}{\columnwidth}{X r r r}
    & \,\,{\bf RT noscat} & \,\,{\bf RT scat} & \,\,{\bf M1} \\
    \hline
    $L_{\nu_e}\,[10^{52}\,{\rm erg}\,{\rm s}^{-1}]$      & 5.95 & 6.01 & 5 \\
    $L_{\bar{\nu}_e}$                                    & 10.6 & 11.8 & 12 \\
    $L_{\nu_x}$                                          & 30.2 & 12.6 & 12 \\
    $R_{\nu_e}-R_{\bar{\nu}_e}\,[10^{57}\,{\rm s}^{-1}]$ & -1.06 & -1.86 & -2 \\
    $\langle \varepsilon_{\nu_e} \rangle\,[{\rm MeV}]$   & 12.08 & 11.88 & 12 \\
    $\langle \varepsilon_{\bar{\nu}_e} \rangle$          & 15.96 & 14.69 & 15 \\
    $\langle \varepsilon_{\nu_x} \rangle$                & 36.25 & 23.67 & 26 \\
    \hline
  \end{tabularx}
\end{table}

@@

\subsubsection*{Acknowledgements}
This research is supported in part by the Joint Institute for Nuclear
Astrophysics---Center for the Evolution of the Elements (JINA-CEE)
and the National Science Fondation under grant numbers X. \todo{grant nos?}
We thank Sherwood Richers for providing his neutrino transport code
\lstinline{Sedonu}~\footnote{\url{https://bitbucket.org/srichers/sedonu}}
for comparisons,
Fran\c{c}ois H\'{e}bert for his timely help in finding and
correcting bugs in the time-interpolation algorithm,
and Luke Roberts for helpful discussions.

>>>>

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter
representative of the fluid in the Massive Collapsed Star test
(Sec.~\ref{ssec:test_collapse}) at the radius 50~km.
For comparision with that test we use the LS180 equation of state
\cite{latt1991-nuc_eos}
\footnote{available at \lstinline{stellarcollapse.org}}
in which the equilibrium $\nu_e$ neutrino chemical potential is
$\eta_{\nu_e}=-0.1555$
(with $\eta_{\bar{\nu}_e}=-\eta_{\nu_e}$ and $\eta_{\nu_x}=0$).
The opacity table is computed using \lstinline{NuLib} from cross-sections
neglecting weak magnetism
\todo{anything else?}
and is stored on a grid covering energy, density, temperature, and composition
ranges spanning $\varepsilon_i\in[1,280.5]$~MeV logarithmically,
$\varrho\in[10^6,6.3\times10^15]\,{\rm g}\,{\rm cm}^{-3}$ logarithmically,
$T\in[0.05,200]$~MeV logarithmically, and 
$Y_e\in[0.035,0.55]$ linearly,
with grid extents$\{18,82,65,51\}$ respectively.
Interpolation is performed linearly.

@@

Expanding the exponential in Eqn.~\ref{eqn:fae_homogeneous_general},
we see the distribution function takes the limiting values:
\begin{equation}
  \label{eqn:fae_homogeneous_limits}
  f_{\rm AE}(\cos A) = f^{\rm eq}
  \begin{cases}
    \chi s(\cos A) & s \ll \chi^{-1}\,\textrm{optically thin}\\
    1                  & s \gg \chi^{-1}\,\textrm{optically thick}.
  \end{cases}
\end{equation}
This agrees with Fig.~\ref{fig:f_absorption_sphere}, with $\nu_x$
in the optically thin limit and $\nu_e$ in the optically thick limit.

@@

\subsection{Post-Bounce Collapse Profile:
  Testing Scattering}
\label{ssec:test_collapse}
To test our scattering treatments we calculate neutrino fields outside a
collapsed 15~${\rm M}_\odot$ star, 100~ms after core bounce,
comparing ray tracing fields to those from a Monte Carlo transport calculation.
Elastic scattering interior to the shock at $r\approx150$~km
significantly modifies the neutrinos' spectra,
and the extended envelope outside the shock becomes a source of
higher-than-average-energy neutrinos.

The 1D matter profile and M1 transport evolution are computed using
open source supernova evolution code \lstinline{GR1D}
\footnote{\url{http://www.gr1dcode.org}}
\cite{ocon2010-gr1d, ocon2015-gr1d_with_nu},
using a progenitor profile from \cite{woos1995-sn_progenitors}.
The matter is described by the LS180 equation of state \cite{latt1991-nuc_eos},
and the opacities are computed and stored in a table
as described in Sec.~\ref{ssec:test_equilibrium}.
This standard test is also presented for example in
\cite{ocon2015-gr1d_with_nu,fouc2015-m1_nsbh,abdi2012-monte_carlo}.
\todo{also lieb2005, mull2010}

The matter profile and background field are stored on a spherical
pseudo-spectral grid composed of 11 spherical-shell subdomains
\todo{cite early SpEC paper}
comprising a total of 62 radial grid points spaced approximately
logarithmically across $r\in(0,740)$~km.
The background scattering field $\Phi(\varepsilon)$ is supplied by
\lstinline{GR1D} in the form of $J(\varepsilon_i)$,
with $\varepsilon_i$ representing 18 energy groups
identical to those in the \lstinline{NuLib} table described in
Sec.~\ref{ssec:test_equilibrium}.
For the \emph{spectral} method we use $J(\varepsilon)$ directly,
using zeroth-order interpolation between energy groups;
for the \emph{gray} method we use $J=\sum J(\varepsilon_i) \Delta\varepsilon_i$,
and $G=\sum J(\varepsilon_i) \Delta\varepsilon_i/\varepsilon_i$ instead.

For fiducial neutrino distributions we use the matter profile
as input into a Monte Carlo radiation transport calculation using
open source neutrino transport code \lstinline{Sedonu}
\footnote{\url{https://bitbucket.org/srichers/sedonu}}
\cite{rich2015-monte_carlo}.
To homogenize the physics modeled across these three treatments
(M1 transport to provide the background fields,
ray tracing to compute neutrino distributions, and
Monte Carlo transport for a fiducial comparison)
we turn off inelastic scattering in the M1 and Monte Carlo codes,
since we have not incorporated that physics into this ray tracing code,
and we turn off general relativistic effects in the ray tracing code,
since the Monte Carlo code assumes Minkowski spacetime.

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-collapse_profile_avg_eps_vs_cosA}
  % Data:   $DIR/
  % Script: $DIR/plot-compare_spectral_vs_gray-angular_distribution_avg_eps-production.gnu
  % w/  DIR_MC     =$DIR/ResultsSherwood-03/Spectra
  %     DIR_SPEC_SC=$DIR/ResultsBrett-A-03/Test-scat-01
  %     DIR_GRAY_SC=$DIR/ResultsBrett-B-03/Test-scat-01
  %     DIR_GRAY_AB=$DIR/ResultsBrett-B-03/Test-noscat-01
  % and DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20161007-ray_tracing_tests/CollapseProfile/
  \caption{Angular distribution of $\nu_x$ average energies
    $\langle\varepsilon\rangle(\cos A)$
    in the collapse profile test described in Sec.~\ref{ssec:test_collapse}.
    The observer is at 500~km, and the shock at 150~km,
    so that the half-opening angle of the shock is $\cos A\approx0.954$.
    The four methods depicted are 1) a fiducial Monte Carlo calculation,
    ray tracing using the 2) spectral and 3) gray methods to estimate background
    fields for scattering, and 4) ray tracing neglecting scattering.
  }
  \label{fig:avg_eps_collapse}
\end{figure}

We place a stationary observer at $r=500$~km.
Fig.~\ref{fig:avg_eps_collapse} shows the distribution of average $\nu_x$
energies across incoming angles to this observer.
We show $\nu_x$ because they present the largest scattering effects:
they scatter through a thicker atmosphere outside their deep emission surface,
and their hotter spectrum experiences stronger modification due to the
$\varepsilon^2$ dependence of the scattering cross-section.
Against the fiducial Monte Carlo distribution, we show the
\emph{noscat} treatment, and the \emph{scat} treatment using both the
\emph{spectral} method and the \emph{gray} method.

As expected the average energies from both
the scattering envelope ($\cos A \lesssim 0.95$) and
the bright core ($\cos A\gtrsim 0.995 $)
are badly characterized by the \emph{noscat} treatment.
In this case for $\nu_x$ the major effect of elastic scattering
is to decrease the average energy of neutrinos coming from the core
and increase the average energy of neutrinos coming from the envelope.

We don't show it here, but the angular distribution contributing to the
total number density is also strongly affected by elastic scattering.
Without elastic scattering, the central regions emitting 60\%
of the neutrinos for the species $\{\nu_e,\bar{\nu}_e,\nu_x\}$
have scales $R_\nu\approx\{40,35,20\}$~km;
with elastic scattering the scales are
$R_\nu\approx\{50,45,35\}$~km.
% From by-hand summation over histogram G(cosA_i)*dcosA
% Script: $DIR/plot-compare_scat_vs_noscat-G_angular_distribution-all.gnu
% DIR=.../20161007-ray_tracing_tests/CollapseProfile/ResultsBrett-B-03
Total luminosities and average energies from the different treatments are
\todo{update MC results 02->03}
presented in Tab.~\ref{tab:collapse_rt_vs_mc}.

\begin{table}%[h]
  % to get mc luminosities: tail $MCDATA/Spectra/luminosities_and_avg_eps-501km-nu?.dat
  % to get rt luminosities: tail $RTDATA/Moments-nu?.out
  % MCDATA=$DIR/ResultsSherwood-02
  % RTDATA_NO=$DIR/ResultsBrett-B-03/Test-noscat-01/
  % RTDATA_GR=$DIR/ResultsBrett-B-03/Test-scat-01/
  % RTDATA_SP=$DIR/ResultsBrett-A-03/Test-scat-01/
  \caption{
    Comparison of total luminosities and average energies between
    the following methods:
    ray tracing ignoring elastic scattering `RT noscat',
    ray tracing with the gray scattering treatment `RT scat gr',
    ray tracing with the spectral scattering treatment `RT scat sp', and
    a fiducial Monte Carlo transport evolution `MC'.
    We also show average energies of the scattering envelope
    $\langle\varepsilon\rangle^{\rm sc}$, which are estimated by eye
    from plots like Fig.~\ref{fig:avg_eps_collapse} at $\cos A < 0.95$.
    Luminosities have units $10^{52}\,{\rm erg}\,{\rm s}^{-1}$ and
    average energies have units MeV.
    The $\nu_x$ luminosities are per-species:
    multiply by four to get the total heavy-lepton neutrino luminosities.
  }
  \label{tab:collapse_rt_vs_mc}
  \begin{tabularx}{\columnwidth}{X r r r r}
    & {\bf RT noscat} & \,\,{\bf RT scat gr} & \,\,{\bf RT scat sp} & \,\,{\bf MC} \\
    \hline
    $L_{\nu_e}$                                         &  3.53 & 3.14 & 3.04 & 3.48 \\
    $L_{\bar{\nu}_e}$                                   & 5.19 & 3.05 & 3.03 & 3.01 \\
    $L_{\nu_x}$                                         & 222.0  & 1.88 & 1.76 & 1.70 \\
    $\langle \varepsilon_{\nu_e} \rangle$               &  10.6 & 10.6 & 10.9 & 11.0 \\
    $\langle \varepsilon_{\bar{\nu}_e} \rangle$         & 14.2 & 13.0 & 13.8 & 13.7 \\
    $\langle \varepsilon_{\nu_x} \rangle$               & 47.8 & 16.0 & 17.3 & 16.2 \\
    $\langle \varepsilon_{\nu_e} \rangle^{\rm sc}$      & 4    & 14   & 17   & 16 \\
    $\langle \varepsilon_{\bar{\nu}_e} \rangle^{\rm sc}$& 4    & 24   & 20   & 18 \\
    $\langle \varepsilon_{\nu_x} \rangle^{\rm sc}$      & 2    & 28   & 31   & 27 \\
    \hline
  \end{tabularx}
\end{table}

We can make some quantitative sense of the scattered energies in
Fig.~\ref{fig:avg_eps_collapse} from the formal solution explored in the
simple configurations above.
In particular, the average energy in the scattering envelope
($\cos A\lesssim 0.95$) is related to the spectrum in the direction
of the interior ($\cos A \sim 1$).
The relation is quite simple if we simplify our realistic model
to that of a homogeneous matter profile and
background scattering field.

In the envelope the local temperature is low
and the ratio $\chi^*_a/\chi_s$ small,
so $f_{\rm SE}$ strongly dominates over $f_{\rm AE}$.
By expanding the exponential of
Eqn.~\ref{eqn:fse_homogeneous_general} as we did for
Eqn.~\ref{eqn:fae_homogeneous_limits},
and furthermore factoring out the dominant ${\rm energy}^2$ dependence from the
scattering opacity $\chi_s\equiv\varepsilon^2 \zeta_s$, we have
\begin{equation}
  \label{eqn:fse_homogeneous_thin_limit}
  f_{\rm SE}(\varepsilon)
  = \varepsilon^2 \zeta_s \,\Phi(\varepsilon)\,s,
\end{equation}
where $s$ is the length of scattering envelope passed through by the ray.
The average energy of the scattered field is
\begin{equation}
  \label{eqn:avg_eps_sc}
  \langle \varepsilon \rangle^{\rm sc}
  = \frac{\int d\varepsilon\,\varepsilon^3 f}{\int d\varepsilon\,\varepsilon^2 f}
  = \frac{\int d\varepsilon\,\varepsilon^5 \Phi}{\int d\varepsilon\,\varepsilon^4 \Phi}.
\end{equation}

The spectrum of the background field $\Phi(\varepsilon)$ is well-approximated
as the \emph{scat} solution for $\cos A\sim 1$, since that is the dominant
source direction for neutrinos.
And because the scattering envelope is optically thin at all energies,
we can assume the neutrino spectrum is essentially unchanged in its passage
through the envelope.
In order to estimate $\langle \varepsilon \rangle^{\rm sc}$ analytically, we may
approximate the background field $\Phi(\varepsilon)$ analytically.
Direct Fermi-Dirac fits using a temperature and chemical potential
representative of a physical neutrinosurface fair poorly since neutrinos of
different energies decouple from the matter at different radii, over which
the thermodynamic state varies substantially.
Phenomenological Fermi-Dirac fits work well;
but so do pinched spectral fits which are much simpler \cite{miri????}:
\begin{equation}
  \label{eqn:pinched_spectra}
  \Phi^{\rm pi}(\varepsilon) \propto
  \varepsilon^{\alpha-2}
  \exp\left(-(\alpha+1)\frac{\varepsilon}{\langle\varepsilon\rangle}\right),
\end{equation}
(where our definition differs from that of \cite{miri????} by a factor of
$\varepsilon^2$, since we define our distribution function to be dimensionless
in natural units $\{\hbar,c\}=1$).

Pinching parameters (calculated by eye) for the species
$\{\nu_e,\bar{\nu}_e,\nu_x\}$ are $\alpha\approx\{3.6,5.1,2.3\}$,
when we use the average energies from the \emph{spectral}
\emph{scat} method in Tab.~\ref{tab:nsns_rt_vs_m1}.
Energy moments of pinched spectra like those in Eqn.~\ref{eqn:avg_eps_sc}
have a simple analytic form so that Eqn.~\ref{eqn:avg_eps_sc} becomes
\begin{align}
  \label{eqn:pinched_avg_eps_sc_1}
  \langle\varepsilon\rangle^{\rm sc}
  &\approx \frac{\alpha+3}{\alpha+1} \langle\varepsilon\rangle, \\
  %\label{eqn:pinched_avg_eps_sc_2}
  %&= \{1.4,1.3,1.6\} \langle\varepsilon\rangle, \\
  \label{eqn:pinched_avg_eps_sc_3}
  &\approx \{16,18,28\}\,~{\rm MeV},
\end{align}
again for $\{\nu_e,\bar{\nu}_e,\nu_x\}$ respectively.
These analytic predictions agree with the average energies of the scattering
envelope to within 20\%
% From by-eye examinations of <eps>(cosA) plot using data from
% MCDATA=$DIR/ResultsSherwood-02
% RTDATA_GR=$DIR/ResultsBrett-B-03/Test-scat-01/
% RTDATA_SP=$DIR/ResultsBrett-A-03/Test-scat-01/
%          nue     nua     nux
of the Monte Carlo and the \emph{scat} ray tracing calculations,
even though we have made some drastic simplifications to our model.

@@

\begin{table}%[h]
  \caption{
    Comparison of total luminosities and average energies between
    ray tracing using the \emph{noscat} method `RT noscat',
    ray tracing using the \emph{gray} \emph{scat} method `RT scat',
    and the M1 transport simulation `M1'.
    The ray tracing totals were computed from sums over
    observers placed in the $y$-$z$ plane along an arc at coordinate distance
    $r=250\,{\rm km}$ from the center of the star, at 30 positions
    distributed evenly in $\cos\theta$ over the colatitude range
    $\theta\in[0,90^{\circ}]$.
    M1 values are taken from \cite[Figs. 7, 9, 10]{fouc2016-m1_evolve_n}.
    Energy luminosities have units $10^{52}\,{\rm erg}\,{\rm s}^{-1}$,
    number luminosities $10^{57}\,{\rm s}^{-1}$, and
    average energies MeV.
    The $\nu_x$ luminosities are per-species:
    multiply by four to get the total heavy-lepton neutrino luminosities.
  }
  \label{tab:nsns_rt_vs_m1}
  \begin{tabularx}{\columnwidth}{X r r r}
    & {\bf RT noscat} & \,\,{\bf RT scat} & \,\,{\bf M1} \\
    \hline
    $L_{\nu_e}$                                 & 5.95 & 6.01 & 5 \\
    $L_{\bar{\nu}_e}$                           & 10.6 & 11.8 & 12 \\
    $L_{\nu_x}$                                 & 30.2 & 12.6 & 12 \\
    $R_{\nu_e}-R_{\bar{\nu}_e}$                 & -1.06 & -1.86 & -2 \\
    $\langle \varepsilon_{\nu_e} \rangle$       & 12.08 & 11.88 & 12 \\
    $\langle \varepsilon_{\bar{\nu}_e} \rangle$ & 15.96 & 14.69 & 15 \\
    $\langle \varepsilon_{\nu_x} \rangle$       & 36.25 & 23.67 & 26 \\
    \hline
  \end{tabularx}
\end{table}

@@

\subsubsection*{Acknowledgements}
This research is supported in part by the Joint Institute for Nuclear
Astrophysics---Center for the Evolution of the Elements (JINA-CEE)
and the National Science Fondation under grant numbers X. \todo{grant nos?}
We thank Sherwood Richers for providing his neutrino transport code
\lstinline{Sedonu} for comparisons,
Fran\c{c}ois H\'{e}bert for his timely help in finding and
correcting bugs in the time-interpolation algorithm,
and Luke Roberts for helpful discussions.
